\vspace{-3mm}
\section{Experiment}
\label{sec:evaluation}

% 减少公式之间的间距
\begingroup
\setlength{\abovedisplayskip}{3pt}
\setlength{\belowdisplayskip}{3pt}
\setlength{\abovedisplayshortskip}{2pt}
\setlength{\belowdisplayshortskip}{2pt}
\setlength{\jot}{2pt}

% EX VES CR AR@1 AR@0.8
% Baselines:
% -  开源基础模型 + 闭源基础模型： GPT-5.4、Deepseek、Qwen-7B、XiYanSQL-7B、Llama-8B (选3-4个模型)
% -  Text-to-SQL Agentic / Approach : 
		%Multi-Agent Approach： Agentar-Scale-SQL、MAC-SQL、CHESS
%		%Single-Agent Appraoch:  DIN-SQL、DAIL-SQL
% Benchmark: BIRD and EESQLBench

% RQ1： Overall Performance  表格可以参考 LEAF-SQL: Level-wise Exploration with Adaptive Fine-graining for Text-to-SQL Skeleton Prediction、SchemaRAG
% RQ2： Ablation Study
% RQ3:  Case Study


%- BaseLines
%1. （2种）基于 Prompt 的方法：DIN-SQL，DAIL-SQL
%2. （3种）基于 Agent 的方法：
%- MAC-SQL，DeepEye
%- CHESS（需要再看一下结果）
%3. （2种）基于微调模型的方法（基于原始 BIRD pipeline）：XiYan-SQL（QwenCoder-32B-2504） + OmniSQL(32B)
%
%- LLMs
%- Qwen3-Coder-30B-A3B-Instruct
%- CodeLlama-34b-Instruct-hf
%- DeepSeek-v4-pro
%- GPT-5.4


% Performance Evaluation  (Token 消耗)
% Ablation Study
% Error Analysis

% 统计显著性检验
% 重复运行5次?


\vspace{-2mm}
\subsection{Experimental Setup}
\textbf{Datasets.}
We evaluate \toolname\ on two complementary benchmarks: \textbf{\textsc{BIRD}}~\cite{li2023can} and \textbf{\textsc{EESQLBench}}~\cite{lin2026cracking}.
%\begin{itemize}[leftmargin=10pt]
\textsc{BIRD} is a widely used cross-domain Text-to-SQL benchmark, containing more than 12,751 question--SQL pairs over 95 large-scale databases and 37 professional domains.
It features realistic database scenarios with complex schemas, domain-specific knowledge, and noisy data, making it suitable for evaluating the general Text-to-SQL capability of \toolname.
%
To further evaluate the generality and efficiency-oriented capability of \toolname, we use \textsc{EESQLBench}, an efficiency-focused benchmark where each NL question is paired with an expert-optimized canonical SQL query.
% containing 213 tasks,
Its tasks include large databases, complex joins, nested subqueries, and optimization-sensitive predicates, enabling us to assess whether \toolname\ can generate SQL queries that are not only semantically correct but also close to human expert-level efficiency.
%\end{itemize}

\textbf{Metrics.}
We evaluate both semantic correctness and execution efficiency.
Let $q_i^*$ be the canonical SQL, $\hat{q}_i$ be the generated SQL, $S_i$ be the database, and $N$ be the number of test instances.
We define:
\[
\mathbf{1}_i =
\begin{cases}
	1, & \mathrm{Exec}(\hat{q}_i,S_i)=\mathrm{Exec}(q_i^*,S_i),\\
	0, & \text{otherwise}.
\end{cases}
\]

\textbf{\textit{Execution Accuracy (EX)}}~\cite{li2023can} measures the fraction of semantically correct outputs:
\[
\mathrm{EX}=\frac{1}{N}\sum_{i=1}^{N}\mathbf{1}_i.
\]
It evaluates whether the generated SQL produces the correct result, regardless of efficiency.

\textbf{\textit{Valid Efficiency Score (VES)}}~\cite{li2023can} measures runtime-based efficiency for valid outputs:
\[
\mathrm{VES}=
\frac{1}{N}\sum_{i=1}^{N}
\mathbf{1}_i
\cdot
\sqrt{\frac{T(q_i^*,S_i)}{T(\hat{q}_i,S_i)}} ,
\]
where $T(q,S)$ denotes execution time.
VES rewards generated SQLs that are both correct and fast, but it may be affected by runtime noise.

\textbf{\textit{Cost Reachability (CR)}}~\cite{lin2026cracking} measures how close a generated SQL is to the expert-optimized SQL in terms of plan-level cost.
For each semantically correct instance, we define the instance-level cost ratio:
\[
\mathrm{CR}_i=
\frac{\mathrm{Cost}(q_i^*,S_i)}
{\mathrm{Cost}(\hat{q}_i,S_i)}.
\]
The overall CR is computed as:
\[
\mathrm{CR}=
\frac{1}{\sum_{i=1}^{N} \mathbf{1}_i}
\sum_{i=1}^{N}
\mathbf{1}_i
\cdot
\sqrt{
	\mathrm{CR}_i
}.
\]
On \textsc{EESQLBench}, $\mathrm{Cost}(\cdot)$ is measured by Total Scanned Rows, i.e., the total number of rows processed by physical operators in the query plan.
A higher CR indicates that the generated SQL is closer to expert-level efficiency.


\textbf{\textit{Acceptable Reachability at $k$ (AR@$k$)}}~\cite{lin2026cracking} measures the fraction of all test instances where the generated SQL is both correct and sufficiently efficient:
\[
\mathrm{AR}@k=
\frac{1}{N}
\sum_{i=1}^{N}
\mathbf{1}_i\cdot \mathbf{1}(\mathrm{CR}_i\ge k).
\]
We report AR@1 and AR@0.8.
AR@1 counts correct queries whose cost is no higher than the expert SQL, while AR@0.8 allows the generated SQL to be at most $1.25\times$ the expert cost.

\textbf{Baselines.}
We compare \toolname\ with three categories of representative Text-to-SQL methods:

\begin{itemize}[leftmargin=10pt]
\item \textbf{\textit{Fine-tuning-based methods.}}
This category includes models that rely on in-domain training data or task-specific fine-tuning to improve Text-to-SQL performance.
We include \textbf{\textsc{XiYan-SQL}}~\cite{liu2025xiyansql} and \textbf{\textsc{Omni-SQL}}~\cite{li2025omnisql}, using their released checkpoints \textsc{XiYanSQL-QwenCoder-32B-2504} and \textsc{OmniSQL-32B}.
These baselines are used to examine whether \toolname, without task-specific fine-tuning, can achieve competitive performance compared with trained Text-to-SQL models.

\item \textbf{\textit{Prompting-based methods.}}
This category contains training-free methods that adapt general-purpose LLMs to Text-to-SQL at inference time through prompting, schema serialization, demonstration retrieval, and reasoning instructions.
We include \textbf{\textsc{DIN-SQL}}~\cite{pourreza2023din} and \textbf{\textsc{DAIL-SQL}}~\cite{gao2023text}.

\item \textbf{\textit{Agentic Text-to-SQL workflows.}}
We further compare with agent-based Text-to-SQL systems, including \textbf{\textsc{MAC-SQL}}~\cite{wang2023mac} and \textbf{\textsc{DeepEye}}~\cite{li2026deepeye}.
These methods decompose Text-to-SQL into multiple agentic steps, such as schema linking, SQL generation, execution-based refinement, and candidate selection.
They provide strong baselines for evaluating whether the meta-cognitive controller and optimization loop in \toolname\ bring additional benefits beyond existing agentic workflows.
\end{itemize}

To ensure a fair comparison, for prompting-based methods and agentic workflows, we evaluate all systems under the same backbone LLMs.
Specifically, we use three representative backbones: \textsc{Qwen3-Coder-30B-A3B-Instruct}, \textsc{DeepSeek-v4-pro}, and \textsc{GPT-5.4}.
%For fine-tuning-based methods, since their performance depends on released task-specific checkpoints, we directly use their official checkpoints and report them separately from the backbone-controlled comparison.


\textbf{Environment.} 
We perform our experiments on a workstation with an 8-core 11th Gen Intel(R) Core(TM) i7-11700 @ 2.50GHz processor, 64GB of memory, and NVIDIA RTX A6000 with 48GB of VRAM running Ubuntu 22.04.1 LTS.
For open-source LLMs, we deploy a local API server based on vLLM~\cite{kwon2023efficient} which is a unified library for LLM serving and inference.
For GPT-5.4, we query the respective official APIs with matching decoding settings.

\subsection{Performance Comparison}
% CR 越高， VES 也越高
% EX >= AR@0.8 >= AR@1

% 总体性能 + 统计验证
\textbf{Overall Results.}
Table~\ref{tab:overall-performance} reports the overall performance of \toolname\ and representative baselines on \textsc{BIRD} and \textsc{EESQLBench}.
EX directly measures how often the generated SQL returns the correct result, while VES can be viewed as a runtime-aware accuracy score: incorrect SQL queries contribute zero, and correct SQL queries receive higher scores when they run faster than the canonical SQL.
CR focuses on the semantically correct queries only, measuring whether their plan-level cost is close to or lower than the canonical SQL.
AR@0.8 and AR@1 further measure the end-to-end fraction of test cases where the generated SQL is both correct and sufficiently efficient.
Overall, \toolname\ achieves the strongest or highly competitive results across both correctness-oriented and efficiency-oriented metrics.
On \textsc{BIRD}, \toolname\ achieves the best EX of 73.29\%, the best VES of 83.02, and the best AR@0.8 and AR@1 of 69.85\% and 67.96\%, respectively.
On \textsc{EESQLBench}, \toolname\ achieves the best EX of 67.89\%, the best VES of 107.16, the highest CR of 13.05, and the best AR@0.8 and AR@1 of 54.19\% and 50.59\%, respectively.
These results show that \toolname\ improves not only SQL correctness but also efficiency.
%We also note that VES and CR capture complementary aspects of efficiency: VES reflects actual runtime behavior, while CR measures plan-level cost reachability.
%Therefore, they may not always move identically due to runtime noise, caching effects, or the gap between scanned rows and actual execution time.

\textbf{Controlled Comparison under Same Backbones.}
We further compare methods under the same backbone LLMs to examine whether the improvements come from the workflow design rather than the underlying model.
On \textsc{BIRD}, 
compared with the strongest agentic baseline \textsc{DeepEye}, \toolname\ improves EX by 0.79, 1.99, and 1.53 percentage points under \textsc{Qwen3-Coder}, \textsc{DeepSeek-v4}, and \textsc{GPT-5.4}, respectively.
It also improves AR@1 by 0.66, 0.67, and 1.28 percentage points under the same three backbones.
Compared with prompting-based methods such as \textsc{DIN-SQL} and \textsc{DAIL-SQL}, \toolname\ achieves substantially higher EX and AR@1, showing stronger end-to-end correctness and efficiency reachability.
Moreover, compared with fine-tuning-based methods such as \textsc{XiYan-SQL} and \textsc{Omni-SQL}, \toolname\ achieves better overall performance without task-specific fine-tuning, improving the best fine-tuned baseline by 14.68 percentage points in EX and 11.31 percentage points in AR@1.
%
%\textbf{Performance on \textsc{EESQLBench}.}
On the more complex and efficiency-oriented \textsc{EESQLBench}, the advantage of \toolname\ becomes more pronounced.
Under the same backbone LLMs, \toolname\ improves EX over \textsc{DeepEye} by 2.53, 1.12, and 3.39 percentage points, and improves AR@1 by 4.40, 1.98, and 0.58 percentage points under \textsc{Qwen3-Coder}, \textsc{DeepSeek-v4}, and \textsc{GPT-5.4}, respectively.
The gains are larger on efficiency-oriented metrics, where \toolname\ improves VES by 18.05, 3.61, and 38.49 points over \textsc{DeepEye} under the same backbones.
These results show that \toolname\ is robust across different backbone LLMs and significantly outperforms prompting-based, fine-tuning-based, and agentic baselines in both correctness and efficiency-oriented performance.

\textbf{Statistical Significance.}
We conduct paired analysis between \toolname\ and the strongest baseline under the same backbone LLMs.
For EX and AR@1, we use McNemar's test over paired success outcomes; for VES and CR score, we use the Wilcoxon signed-rank test over paired scores.
The main comparisons are statistically significant ($p < 0.001$), and the 95\% bootstrap confidence intervals are consistently positive, with a medium effect size of approximately $\delta=0.52$.

\textbf{Token Consumption.}
We compare the average token consumption per task on \textsc{EESQLBench} under \textsc{Qwen3-Coder}.
\toolname\ consumes only 613K tokens per task on average, reducing token usage by 63.7\% over \textsc{MAC-SQL} and 89.8\% over \textsc{DeepEye}.
This shows the advantage of adaptive workflow control: \toolname\ avoids applying expensive reasoning steps uniformly, and instead allocates evidence construction, reflection, and optimization only when they are needed.

% 保留两位小数
\begin{table*}[t]
	\centering
	\vspace{-3mm}
	\caption{\textbf{Overall performance comparison on \textsc{BIRD} and \textsc{EESQLBench}.}
		For both benchmarks, we report EX, VES, CR, AR@0.8, and AR@1.
		The best results are highlighted in bold.}
	\vspace{-2mm}
	\label{tab:overall-performance}
	\footnotesize
	\setlength{\tabcolsep}{6.0pt}
	\renewcommand{\arraystretch}{0.85}
	\begin{tabular}{llcccccccccc}
		\toprule
		\multirow{2}{*}{\textbf{Method}}
		& \multirow{2}{*}{\textbf{LLM}}
		& \multicolumn{5}{c}{\textbf{\textsc{BIRD}}}
		& \multicolumn{5}{c}{\textbf{\textsc{EESQLBench}}} \\
		\cmidrule(lr){3-7}
		\cmidrule(lr){8-12}
		& & \textbf{EX(\%)} & \textbf{VES} & \textbf{CR} & \textbf{AR@0.8(\%)} & \textbf{AR@1(\%)}
		& \textbf{EX(\%)} & \textbf{VES} & \textbf{CR} & \textbf{AR@0.8(\%)} & \textbf{AR@1(\%)} \\
		\midrule
		
		\multicolumn{12}{l}{\textit{Fine-tuning-based methods}} \\
		\midrule
		\textsc{XiYan-SQL} 
		& \textsc{XiYan-32B} 
		& 58.61 & 71.28 & 1.01 & 57.37 & 56.65 & 16.13 & 25.24 & 3.35 & 16.12 & 16.12 \\
		
		\textsc{Omni-SQL} 
		& \textsc{OmniSQL-32B} 
		& 49.41 & 57.03 & 1.00 & 47.45 & 46.48 & 15.05 & 20.87 & 1.44 & 14.52 & 14.52 \\
		
		\midrule
		\multicolumn{12}{l}{\textit{Prompting-based methods}} \\
		\midrule
		\multirow{3}{*}{\textsc{DIN-SQL}}
		& \textsc{Qwen3-Coder} 
		& 57.63 & 64.29 & 1.86 & 54.41 & 52.80 & 18.29 & 47.63 & 5.81 & 13.50 & 12.53 \\
		& \textsc{DeepSeek-v4} 
		& 59.06 & 66.43 & 1.77 & 56.30 & 55.00 & 40.85 & 31.15 & 6.86 & 39.80 & 29.41 \\
		& \textsc{GPT-5.4} 
		& 42.05 & 46.42 & 1.42 & 39.81 & 38.62 & 49.39 & 67.44 & 10.92 & 38.16 & 36.81 \\
		\midrule
		
		\multirow{3}{*}{\textsc{DAIL-SQL}}
		& \textsc{Qwen3-Coder} 
		& 60.63 & 66.32 & 1.50 & 56.01 & 54.31 & 43.90 & 53.39 & 4.29 & 32.70 & 31.52 \\
		& \textsc{DeepSeek-v4} 
		& 39.99 & 65.64 & \textbf{2.25} & 38.73 & 37.81 & 54.88 & 47.78 & 7.62 & 38.50 & 37.12 \\
		& \textsc{GPT-5.4} 
		& 62.26 & 68.61 & 1.70 & 59.51 & 57.50 & 57.32 & 69.11 & 10.64 & 40.62 & 39.91 \\
		
		\midrule
		\multicolumn{12}{l}{\textit{Agentic Text-to-SQL workflows}} \\
		\midrule
		\multirow{3}{*}{\textsc{MAC-SQL}}
		& \textsc{Qwen3-Coder} 
		& 65.71 & 72.21 & 1.83 & 61.40 & 59.12 & 57.32 & 51.68 & 8.04 & 39.10 & 37.51 \\
		& \textsc{DeepSeek-v4} 
		& 63.89 & 74.14 & 2.06 & 57.42 & 55.60 & 59.15 & 49.01 & 6.86 & 40.22 & 38.62 \\
		& \textsc{GPT-5.4} 
		& 66.56 & 75.47 & 1.87 & 59.60 & 57.12 & 60.37 & 49.92 & 8.49 & 40.82 & 36.20 \\
		\midrule
		
		\multirow{3}{*}{\textsc{DeepEye}}
		& \textsc{Qwen3-Coder} 
		& 72.50 & 81.52 & 1.69 & 69.60 & 67.30 & 59.76 & 62.26 & 7.37 & 46.70 & 46.19 \\
		& \textsc{DeepSeek-v4} 
		& 69.56 & 77.35 & 1.69 & 65.48 & 63.24 & 62.13 & 56.55 & 6.94 & 47.30 & 46.12 \\
		& \textsc{GPT-5.4} 
		& 68.45 & 74.94 & 1.70 & 63.41 & 61.30 & 64.50 &68.67 & 3.96 & 46.30 & 43.31 \\
		
		\midrule
		\rowcolor{gray!10}
		& \textsc{Qwen3-Coder}
		& \textbf{73.29} & 82.42 & 1.70 & \textbf{69.85} & \textbf{67.96}
		& 62.29 & 80.31 & \textbf{13.05} & 51.42 & \textbf{50.59} \\
		
		\rowcolor{gray!10}
		& \textsc{DeepSeek-v4}
		& 71.55 & \textbf{83.02} & 1.74 & 66.05 & 63.91
		& 63.25 & 60.16 & 10.70 & \textbf{54.19} & 48.10 \\
		
		\rowcolor{gray!10}
		\multirow{-3}{*}{\cellcolor{gray!10}\textbf{\toolname}}
		& \textsc{GPT-5.4}
		& 69.98 & 76.32 & 1.72 & 64.85 & 62.58
		& \textbf{67.89} & \textbf{107.16} & 6.76 & 51.62 & 43.89 \\
		
		
		\bottomrule
	\end{tabular}
	\vspace{-4mm}
	
\end{table*}

%\begin{table*}[t]
%	\centering
%	\vspace{-2mm}
%	\caption{\textbf{Overall performance comparison on \textsc{BIRD} and \textsc{EESQLBench}.}
%		For both benchmarks, we report EX, VES, CR, AR@0.8, and AR@1.
%		The best results are highlighted in bold.}
%	\vspace{-2mm}
%	\label{tab:overall-performance}
%	\footnotesize
%	\setlength{\tabcolsep}{6.0pt}
%	\renewcommand{\arraystretch}{0.95}
%	\begin{tabular}{llcccccccccc}
%		\toprule
%		\multirow{2}{*}{\textbf{Method}}
%		& \multirow{2}{*}{\textbf{LLM}}
%		& \multicolumn{5}{c}{\textbf{\textsc{BIRD}}}
%		& \multicolumn{5}{c}{\textbf{\textsc{EESQLBench}}} \\
%		\cmidrule(lr){3-7}
%		\cmidrule(lr){8-12}
%		& & \textbf{EX(\%)} & \textbf{VES} & \textbf{CR} & \textbf{AR@0.8(\%)} & \textbf{AR@1(\%)}
%		& \textbf{EX(\%)} & \textbf{VES} & \textbf{CR} & \textbf{AR@0.8(\%)} & \textbf{AR@1(\%)} \\
%		\midrule
%		
%		\multicolumn{12}{l}{\textit{Fine-tuning-based methods}} \\
%		\midrule
%		\textsc{XiYan-SQL} 
%		& \textsc{XiYan-32B} 
%		& 58.6 & 71.3 & 1.0 & 57.4 & 56.7 & 16.1 & 25.2 & 3.4 & 16.1 & 16.1 \\
%		
%		\textsc{Omni-SQL} 
%		& \textsc{OmniSQL-32B} 
%		& 49.4 & 57.0 & 1.0 & 47.5 & 46.5 & 15.1 & 20.9 & 1.4 & 14.5 & 14.5 \\
%		
%		\midrule
%		\multicolumn{12}{l}{\textit{Prompting-based methods}} \\
%		\midrule
%		\multirow{3}{*}{\textsc{DIN-SQL}}
%		& \textsc{Qwen3-Coder} 
%		& 57.6 & 64.3 & 1.9 & 54.4 & 52.8 & 18.3 & 47.6 & 5.8 & 13.5 & 12.5 \\
%		& \textsc{DeepSeek-v4} 
%		& 59.1 & 66.4 & 1.8 & 56.3 & 55.0 & 40.9 & 31.2 & 6.9 & 39.8 & 29.4 \\
%		& \textsc{GPT-5.4} 
%		& 42.1 & 46.4 & 1.4 & 39.8 & 38.6 & 49.4 & 67.4 & 10.9 & 38.2 & 36.8 \\
%		\midrule
%		
%		\multirow{3}{*}{\textsc{DAIL-SQL}}
%		& \textsc{Qwen3-Coder} 
%		& 60.6 & 66.3 & 1.5 & 56.0 & 54.3 & 43.9 & 53.4 & 4.3 & 32.7 & 31.5 \\
%		& \textsc{DeepSeek-v4} 
%		& 40.0 & 65.6 & \textbf{2.3} & 38.7 & 37.8 & 54.9 & 47.8 & 7.6 & 38.5 & 37.1 \\
%		& \textsc{GPT-5.4} 
%		& 62.3 & 68.6 & 1.7 & 59.5 & 57.5 & 57.3 & 69.1 & 10.6 & 40.6 & 39.9 \\
%		
%		\midrule
%		\multicolumn{12}{l}{\textit{Agentic Text-to-SQL workflows}} \\
%		\midrule
%		\multirow{3}{*}{\textsc{MAC-SQL}}
%		& \textsc{Qwen3-Coder} 
%		& 65.7 & 72.2 & 1.8 & 61.4 & 59.1 & 57.3 & 51.7 & 8.0 & 39.1 & 37.5 \\
%		& \textsc{DeepSeek-v4} 
%		& 63.9 & 74.1 & 2.1 & 57.4 & 55.6 & 59.2 & 49.0 & 6.9 & 40.2 & 38.6 \\
%		& \textsc{GPT-5.4} 
%		& 66.6 & 75.5 & 1.9 & 59.6 & 57.1 & 60.4 & 49.9 & 8.5 & 40.8 & 36.2 \\
%		\midrule
%		
%		\multirow{3}{*}{\textsc{DeepEye}}
%		& \textsc{Qwen3-Coder} 
%		& 72.5 & 81.5 & 1.7 & 69.6 & 67.3 & 59.8 & 62.3 & 7.4 & 46.7 & 46.2 \\
%		& \textsc{DeepSeek-v4} 
%		& 69.6 & 77.4 & 1.7 & 65.5 & 63.2 & 62.1 & 56.6 & 6.9 & 47.3 & 46.1 \\
%		& \textsc{GPT-5.4} 
%		& 68.5 & 74.9 & 1.7 & 63.4 & 61.3 & 64.5 & 68.7 & 4.0 & 46.3 & 43.3 \\
%		
%		\midrule
%		\rowcolor{gray!10}
%		& \textsc{Qwen3-Coder}
%		& \textbf{73.3} & 82.4 & 1.7 & \textbf{69.9} & \textbf{68.0}
%		& 62.3 & 80.3 & \textbf{13.1} & 51.4 & \textbf{50.6} \\
%		
%		\rowcolor{gray!10}
%		& \textsc{DeepSeek-v4}
%		& 71.6 & \textbf{83.0} & 1.7 & 66.1 & 63.9
%		& 63.3 & 60.2 & 10.7 & \textbf{54.2} & 48.1 \\
%		
%		\rowcolor{gray!10}
%		\multirow{-3}{*}{\cellcolor{gray!10}\textbf{\toolname}}
%		& \textsc{GPT-5.4}
%		& 70.0 & 76.3 & 1.7 & 64.9 & 62.6
%		& \textbf{67.9} & \textbf{107.2} & 6.8 & 51.6 & 43.9 \\
%		
%		\bottomrule
%	\end{tabular}
%	\vspace{-4mm}
%\end{table*}

\subsection{Ablation Study}
% 效率评估： 在一个 GPT 上面做就好了
% 评估指标选取 EX （反应correct的代表性指标) 和 AR@1 （反应 correct 和 efficiency 的代表性指标）

%Full OptSQL
%w/o Dynamic Controller  把 Meta-Cognitive Controller 替换成固定 workflow
%w/o Process-level Reflection  只保留最终 validation/repair，不允许中间切换策略、补 evidence、重试 rewrite
%w/o Optimization Loop 生成 initial SQL 后直接输出
%w/o AEKB 不使用历史 rewrite cases，只用当前 query plan 和静态 rules

To understand the contribution of each core component in \toolname, we conduct an ablation study on both \textsc{BIRD} and \textsc{EESQLBench} using GPT-5.4 as the backbone LLM.
We report EX to measure semantic correctness and AR@1 to measure the fraction of generated SQL queries that are both correct and no more expensive than the canonical SQL.
We compare the full \toolname\ with four ablated variants:

\begin{itemize}[leftmargin=10pt]
%	\item \textbf{w/o Adaptation}: replaces the Meta-Cognitive Controller with a fixed workflow.
	\item \textbf{w/o Adaptation}: disables the adaptive strategy-selection ability of the Meta-Cognitive Controller and uses a fixed workflow for all tasks.
	\item \textbf{w/o Reflection}: removes process-level reflection and only keeps final validation and repair.
	\item \textbf{w/o Optimization}: disables the Optimization Loop and directly outputs the SQL after generation phase.
	\item \textbf{w/o AEKB}: disables AEKB-based historical rewrite-case retrieval and only uses current query plans and static optimization rules.
\end{itemize}

\begin{table}[t]
	\centering
%	\vspace{-2mm}
	\caption{\textbf{Ablation study on \textsc{BIRD} and \textsc{EESQLBench}.}
		We use \textsc{GPT-5.4} as the backbone LLM and report EX and AR@1.}
	\vspace{-2mm}
	\label{tab}
	\footnotesize
	\setlength{\tabcolsep}{5.8pt}
	\renewcommand{\arraystretch}{0.95}
	\begin{tabular}{lcccc}
		\toprule
		\multirow{2}{*}{\textbf{Variant}}
		& \multicolumn{2}{c}{\textbf{\textsc{BIRD}}}
		& \multicolumn{2}{c}{\textbf{\textsc{EESQLBench}}} \\
		\cmidrule(lr){2-3}
		\cmidrule(lr){4-5}
		& \textbf{EX(\%)} & \textbf{AR@1(\%)}
		& \textbf{EX(\%)} & \textbf{AR@1(\%)} \\
		\midrule
		Full & 69.98 & 62.58 & 67.89 & 43.89 \\
		w/o Adaptation & 68.45 & 61.82 & 65.24 & 40.34 \\
		w/o Reflection & 67.32 & 58.15 & 58.80 & 36.89 \\
		w/o Optimization & 69.98 & 57.25 & 67.89 & 25.05 \\
		w/o AEKB & 69.98 & 59.60 & 67.89 & 38.65 \\
		\bottomrule
	\end{tabular}
	\vspace{-6mm}
\end{table}

Table~\ref{tab} shows that all core components contribute to the final performance of \toolname.
Removing workflow adaptation and process-level reflection both reduce EX and AR@1, indicating that dynamic strategy selection and intermediate self-reflection are important for handling diverse database environments and complex reasoning paths.
In contrast, disabling the Optimization Loop or AEKB keeps EX unchanged but lowers AR@1, suggesting that these components mainly improve efficiency while preserving semantic correctness.
The large AR@1 drop on \textsc{EESQLBench}, from 43.89\% to 25.05\% without optimization and to 38.65\% without AEKB, confirms the benefit of plan-aware rewriting and reusable optimization experience for efficiency-oriented tasks.

\subsection{Error Analysis}

To better understand the limitations of \toolname, we conduct a qualitative error analysis from both correctness and efficiency perspectives.
Following a standard open-coding protocol, three annotators first independently propose fine-grained labels for recurring error patterns, and then iteratively compare, refine, and merge overlapping labels through discussion.
For \textbf{correctness analysis}, we randomly sample 100 cases judged as incorrect by EX and manually inspect the question, evidence, predicted SQL, and gold SQL.
After resolving disagreements, we consolidate the errors into two high-level groups and 16 fine-grained labels: intent-violating hallucination (87\%), where the predicted SQL changes the user intent, and potential gold error (13\%), where the reference SQL is likely incomplete, ambiguous, or inconsistent with the question.
The most representative fine-grained categories are join logic hallucination (21\%) and DISTINCT error (15\%), suggesting that join-path reasoning and duplicate semantics remain major sources of correctness failures.
%
For \textbf{inefficiency analysis}, we randomly sample 100 semantically correct cases with $\mathrm{CR}_i < 1.0$ and compare the generated SQL, expert SQL, and their execution plans.
Following the same coding protocol, we consolidate recurring inefficiency patterns into five high-level themes and 23 fine-grained labels: join logic (58\%), subquery logic (5\%), timing strategy (5\%), access strategy (4\%), and others (28\%).
The dominant fine-grained causes are redundant lookup joins (48\%) and predicate pull-down opportunities (23\%), indicating that many remaining inefficiencies come from conservative join construction and missed filter-placement rewrites.
We hypothesize that the failures mainly stem from the limited capabilities of individual specialized agents, such as insufficient join-path reasoning and duplicate-semantics analysis.
In future work, we plan to explore more specialized agents with stronger schema grounding, semantic validation, and SQL optimization capabilities.
Due to space limitations, we release the detailed coding results, label definitions, and examples in our GitHub repository~\cite{optsql_error_analysis}.


\endgroup