\section{Motivation}
\label{sec:preliminaries}

\subsection{Task Formulation}
Traditional Text-to-SQL aims to translate a NL question $x$ into an executable SQL query $q$ over a database environment $\mathcal{D}$.
Given a ground-truth query $q^*$, the standard objective is to generate a query whose execution result matches that of $q^*$:
\[
f_\theta(x, \mathcal{D}) \rightarrow q,
\quad
\text{s.t.} \quad
\text{Exec}(q, \mathcal{D}) = \text{Exec}(q^*, \mathcal{D}),
\]
where $f_\theta$ denotes a Text-to-SQL model parameterized by $\theta$, $\mathcal{D}$ denotes the database environment including schema, data, indexes, and statistics, and $\text{Exec}(q, \mathcal{D})$ denotes the result of executing $q$ on $\mathcal{D}$.

However, this formulation focuses mainly on semantic correctness and does not explicitly consider query efficiency.
In practice, multiple semantically equivalent SQL queries may produce the same result but incur substantially different execution costs due to join order, predicate placement, aggregation strategy, or index utilization.
We therefore formulate \textbf{Efficiency-Oriented Text-to-SQL}, where the goal is to generate a query that is both semantically correct and efficient.
Given a NL question $x$ and a database environment $\mathcal{D}$, the goal is to generate a SQL query $q_{\text{opt}}$ satisfying:
\[
\begin{aligned}
	q_{\mathrm{opt}}
	&= \arg\min_{q \in \mathcal{Q}_{\mathrm{corr}}(x,\mathcal{D})}
	\; \text{Cost}(q,\mathcal{D}), \\
	\mathcal{Q}_{\mathrm{corr}}(x,\mathcal{D})
	&= \{q \in \mathcal{Q}(x,\mathcal{D}) \mid \text{Exec}(q,\mathcal{D}) = \text{Exec}(q^*,\mathcal{D})\},
\end{aligned}
\]
where $\mathcal{Q}_{\mathrm{corr}}(x,\mathcal{D})$ denotes the subset of candidates that are semantically correct with respect to $q^*$.
$\text{Cost}(q, \mathcal{S})$ represents the estimated or actual execution cost of query $q$ on $\mathcal{S}$.
Intuitively, the model should not only ensure that $q_{\text{opt}}$ yields correct results but also that it achieves minimal execution time or resource consumption under the same semantic constraints. 

\subsection{Design Objective}
Existing Text-to-SQL systems are typically static pipelines that decompose SQL construction into pre-defined stages, including pre-generation, generation, and post-generation validation. 
While such designs improve robustness compared to single-pass generation, they lack the ability to dynamically adapt their reasoning process to different database environments, query difficulties, and intermediate execution feedback.


\textbf{Beyond Static Pipelines: A Meta-Cognitive Multi-Agent Perspective.}
Inspired by enactive cognition and meta-cognitive theory, we reformulate Text-to-SQL as a dynamic interaction process between an intelligent system and the database environment~\cite{varela1991embodied,flavell1979metacognition}.
Enactive cognition emphasizes that intelligence emerges through continuous interaction with the environment, while meta-cognition enables the system to monitor, regulate, and adapt its own reasoning process.
Based on these principles, we propose a Meta-Cognitive Multi-Agent Text-to-SQL System, where a central Meta-Cognitive Controller dynamically orchestrates a set of specialized agents (actions) under different execution contexts.
Instead of a fixed pipeline, all stages in Text-to-SQL are decomposed into reusable and composable actions, which are invoked online based on the current state of the problem and environment feedback.
This design establishes a unified closed-loop system in which a Meta-Cognitive Controller coordinates planning, execution, evaluation, and optimization over SQL programs based on continuous database feedback.

\textbf{Key Design Principles.}
\textbf{(1) Environment-aware Workflow Adaptation via Strategy Selection.}
The Meta-Cognitive Controller explicitly assesses both the database environment and the complexity of the input query to determine the appropriate reasoning strategy.
For example, for simple queries over well-structured schemas (e.g., single-table filtering), the controller may directly invoke a lightweight generation path using few-shot prompting.
In contrast, for complex analytical queries (e.g., multi-join or domain-specific reasoning), the controller activates a deeper workflow that includes multi-candidate generation, iterative refinement, and optimization loops.
This adaptive strategy selection avoids over- or under-utilization of reasoning resources.

\textbf{(2) Process-level Self-Reflection and Dynamic Control.}
Unlike conventional post-hoc validation approaches, our system enables continuous monitoring of intermediate reasoning states during SQL generation.
The Meta-Cognitive Controller observes signals such as evidence quality, candidate stability, execution behavior, semantic validation results, and optimization progress.
For example, when schema or value evidence is insufficient, the controller may request additional evidence construction through reverse schema linking, value-grounded schema linking, or relational completion.
Similarly, if repeated optimization iterations fail to reduce execution cost, the controller may terminate the current loop and switch to a different reasoning configuration.
This enables closed-loop control over the entire reasoning trajectory.

\textbf{(3) Execution- and Plan-aware Optimization via Feedback-driven Rewriting.}
Beyond generating semantically correct SQL, the system treats execution efficiency as a first-class objective.
Execution results, query plans, and cost estimates are used as feedback signals to reveal whether an initially correct query may still be inefficient.
When such feedback indicates potential performance bottlenecks, the system can refine the SQL through feedback-driven rewriting while preserving its intended semantics.
More importantly, optimization is not treated as an isolated one-shot action.
The system accumulates validated optimization experience from previous interactions with the database environment, enabling future decisions to be guided not only by static rules but also by reusable execution-aware experience.
In this way, SQL optimization becomes a closed-loop process in which database feedback continuously informs rewriting, validation, and future strategy selection.


