\vspace{-1mm}
\section{Related Work}
\label{sec:relatedwork}

%\textbf{Fine-tuned and Trained Text-to-SQL Models.}
%A dominant line of recent work focuses on adapting or training LLMs for Text-to-SQL via supervised fine-tuning, synthetic data construction, or reinforcement learning~\cite{sun2023sqlpalm,li2024codes,pourreza2024dtssql,chen2024opensql,thorpe2024dubo,ma2025sqlr1,liu2025xiyansql,li2025omnisql}.
%For example, \textsc{XiYan-SQL}~\cite{liu2025xiyansql} uses multi-task fine-tuning and a multi-generator ensemble to produce diverse SQL candidates, while \textsc{OmniSQL}~\cite{li2025omnisql} improves model training through large-scale synthesized Text-to-SQL data.
%Reinforcement-learning-based methods such as \textsc{Reasoning-SQL}~\cite{pourreza2025reasoningsql} and \textsc{Arctic-Text2SQL-R1}~\cite{yao2025arctictext2sql} further train models with SQL-tailored rewards or execution-based feedback to improve reasoning and executability.
%Despite their strong performance on benchmarks, these methods remain heavily dependent on training data quality and coverage.
%They often exhibit limited generalization in out-of-domain scenarios and may suffer from reduced robustness when facing complex or real-world database environments~\cite{li2026deepeye}.
%%These methods substantially advance Text-to-SQL by improving the underlying generation model, but their effectiveness still depends on the availability and coverage of training data, the quality of synthesized examples or reward signals, and the benchmark distributions used for training and evaluation.
%
%\textbf{Prompting-based Text-to-SQL Methods.}
%A major line of work adapts general-purpose LLMs to Text-to-SQL through carefully designed prompting strategies, without updating model parameters.
%These methods mainly improve SQL generation by organizing database schemas, sampled contents, task instructions, demonstrations, and decomposition hints into effective prompts~\cite{gao2023text,li2023can,sun2023sql,pourreza2023din}.
%Subsequent studies further enhance prompting with dynamic few-shot selection, self-correction, consistency alignment, and retrieval-augmented evidence construction~\cite{xie2025opensearchsql,ramachandran2024textsqlcalibration,talaei2024chess,pourreza2024chasesql,wang2025linkalign,maamari2024death}.
%These methods are attractive because they are training-free and can leverage increasingly strong backbone LLMs, but their effectiveness often depends on prompt design, retrieved evidence quality, and the capability of the underlying model.
%
%\textbf{Agentic Text-to-SQL Workflows.}
%Recent work further extends LLM prompting into agentic workflows, where LLMs call tools, inspect database context, execute SQL, observe feedback, and revise intermediate outputs. 
%General agent frameworks such as \textsc{ReAct}~\cite{yao2023react}, \textsc{Self-Refine}~\cite{madaan2023selfrefine}, and \textsc{Reflexion}~\cite{shinn2023reflexion} demonstrate the value of combining reasoning, acting, and iterative self-improvement. 
%Following this paradigm, Text-to-SQL agents organize generation around modules such as schema exploration, candidate generation, execution-based repair, multi-path reasoning, semantic memory, and agentic view construction~\cite{wang2023mac,pourreza2024chasesql,deng2025reforce,chaturvedi2025sqlthought,heidari2025agentiql,yang2025marssql,pham2026avsql,biswal2026agentsm}.
%These systems are more robust than prompting-based methods because they can query the database, compare alternatives, and use observed failures to revise SQL.
%As discussed in Section~\ref{sec:intro}, most existing agents still follow largely predefined workflows and strategy choices, use feedback mainly for post-generation repair, validation, or candidate selection, and pay limited attention to efficiency optimization.
%
%\textbf{Difference from \toolname.}
%Existing prompting-based and agentic Text-to-SQL methods mainly focus on improving correctness through better grounding, decomposition, and post-hoc refinement.
%In contrast, \toolname\ is designed as an efficiency- and process-aware framework that goes beyond static or fixed workflows.
%%
%It introduces a meta-cognitive controller that enables (i) environment-aware workflow adaptation, (ii) process-level self-reflection during generation, and (iii) execution- and plan-aware optimization guided by database feedback.
%As a result, \toolname\ treats Text-to-SQL as a dynamic interaction process where both reasoning strategy and optimization behavior are continuously adjusted according to the database environment and intermediate execution signals, rather than following a predefined pipeline or relying only on post-generation correction.

% 精简后的 related work
\textbf{Fine-tuned and Prompting-based Text-to-SQL Methods.}
Fine-tuned methods adapt or train models with supervised fine-tuning, synthetic data, or reinforcement learning~\cite{sun2023sqlpalm,li2024codes,pourreza2024dtssql,chen2024opensql,thorpe2024dubo,ma2025sqlr1,liu2025xiyansql,li2025omnisql,pourreza2025reasoningsql,yao2025arctictext2sql}.
However, these methods depend heavily on training data quality and coverage, and may generalize poorly to out-of-domain or complex real-world databases~\cite{li2026deepeye}.
Prompting-based methods adapt LLMs without updating parameters by organizing schemas, sampled contents, demonstrations, decomposition hints, and retrieved evidence into effective prompts~\cite{gao2023text,li2023can,sun2023sql,pourreza2023din,xie2025opensearchsql,ramachandran2024textsqlcalibration,talaei2024chess,pourreza2024chasesql,wang2025linkalign,maamari2024death}.
Although these methods are training-free, their performance remains sensitive to prompt design, evidence quality, and backbone model capability.

\textbf{Agentic Text-to-SQL Workflows.}
Recent work further extends prompting into agentic workflows, where LLMs inspect database context, call tools, execute SQL, observe feedback, and revise outputs.
General agent frameworks such as \textsc{ReAct}~\cite{yao2023react}, \textsc{Self-Refine}~\cite{madaan2023selfrefine}, and \textsc{Reflexion}~\cite{shinn2023reflexion} show the value of combining reasoning, acting, and iterative refinement.
Following this paradigm, Text-to-SQL agents decompose SQL construction into modules such as schema exploration, candidate generation, execution-based repair, multi-path reasoning, semantic memory, and agentic view construction~\cite{wang2023mac,pourreza2024chasesql,deng2025reforce,chaturvedi2025sqlthought,heidari2025agentiql,yang2025marssql,pham2026avsql,biswal2026agentsm}.
Although these systems are more robust than single-pass prompting, they usually follow largely predefined workflows, use feedback mainly for post-generation repair or candidate selection, and pay limited attention to query efficiency.

\textbf{Difference from \toolname.}
Existing prompting-based and agentic Text-to-SQL methods mainly focus on improving correctness through better grounding, decomposition, and post-hoc refinement.
In contrast, \toolname\ is designed as an efficiency- and process-aware framework that goes beyond static or fixed workflows.
%
It introduces a meta-cognitive controller that enables (i) environment-aware workflow adaptation, (ii) process-level self-reflection during generation, and (iii) execution- and plan-aware optimization guided by database feedback.
As a result, \toolname\ treats Text-to-SQL as a dynamic interaction process where both reasoning strategy and optimization behavior are continuously adjusted according to the database environment and intermediate execution signals, rather than following a predefined pipeline or relying only on post-generation correction.
